Gut microbes and their genes are associated with cognition and neuroanatomy in children

### 2023-03-13 manuscript/main.tex
remove discussion from Figure 2C-G

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N\textsubscript{species} = 92 for 0--120 months, 
N\textsubscript{species} = 46 for 0--6 months, N\textsubscript{species} = 97 for 18--120 months).
Only \emph{Blautia wexlerae} was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B).
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}
\emph{Asaccharobacter celatus} and \emph{Adelcreutzia equolifaciens}
(also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
and \textit{Adlercreutzia equolifaciens} subsp. \textit{equolifaciens} \cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B).

@@

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \cite{dalileRoleShortchainFatty2019}.

We found that synthesis of menaquinone (vitamin K) was negatively associated with
cognitive function score in older children (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183). Menaquinone has several isoforms, one of
which, MK-4, has been found to be decreased in children diagnosed with ASD
\cite{dongCorrelationSerumConcentrations2021}
and is potentially neuroprotective in both rodents and humans
\cite{elkattawyVitaminK2Menaquinone72022}.
In children over 18 months old,
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, q
= 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041). Glutamate is a
critical neurotransmitter controlling neuronal excitatory/inhibitory
signaling along with gamma-aminobutyric acid (GABA), and their balance
in the brain controls neural plasticity and learning, particularly in
the developing brain
\cite{cohenkadoshLinkingGABAGlutamate2015,palomo-buitragoGlutamateInteractionsObesity2019}.
Tryptophan metabolism, including
microbial metabolism of tryptophan, has previously been linked with
autism in children
\cite{hoshinoBloodSerotoninFree1984,xiaoFecalMicrobiomeTransplantation2021}.
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

>>>>

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N\textsubscript{species} = 92 for 0--120 months, 
N\textsubscript{species} = 46 for 0--6 months, N\textsubscript{species} = 97 for 18--120 months).
Only \emph{Blautia wexlerae} was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B).
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae},
\emph{Asaccharobacter celatus} and \emph{Adelcreutzia equolifaciens}
(also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
and \textit{Adlercreutzia equolifaciens} subsp. \textit{equolifaciens} \cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B).

@@

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
In children over 18 months old, synthesis of menaquinone (vitamin K)
was also negatively associated with cognitive function score (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183), while
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, 
q = 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041).
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.
